# Environment {#sec-environment}

Programming necessarily presupposes a programming _environment_. Just as one
can't write text without choosing a font, one can't write code outside of an
environment.

::: callout-note
## Note

Keep in mind that in R, "environment" usually has a very
[specific meaning](https://adv-r.hadley.nz/environments.html). In this chapter
this word refers to the state of your computer while you're programming,
including the organization of your files, the version of R that you're using,
etc.
:::

The environment of an R script is like air is to us: you might forget that it
exists, but it is always there. Then why write a whole chapter about this? Why
write about air?

```{r}
# Printing usually works like this
print(1:9)
```

```{r}
#| echo: false
options(width = 10)
```

```{r}
# What happened here?
print(1:9)
```

```{r}
#| echo: false
options(width = 80)
```

Breathing foul air can make you dizzy, and working in a broken environment might
cause your script to function incorrectly; this is why we need to learn how to
properly manage it. There is no trick in the code above, by the way. I haven't
redefined the `print()` function, I just changed the `width` option in R's
environment.

## File system

In [Chapter -@sec-r-projects], we'll discuss how to organize an R project, but
first we need some guidelines on how to organize files in general. Many people
underestimate the usefulness organizing and naming their files consistently, but
the truth is that having a system saves you valuable time when searching through
your past projects.

First of all, I suggest you create a folder in your computer for all your
programming needs. It's not that rare to create _a lot_ of files when coding,
and making sure they will all be in a single place really helps. Since scripts
are able to write and delete files, this lowers the odds that an R program
ruins your valuable personal files by accident.

Now for the files. Danielle Navarro has an amazing
[presentation](https://slides.djnavarro.net/project-structure) about project
structure where she outlines three main principles to go by when naming files:

1. Be nice to machines.
1. Be nice to humans.
1. Make sorting and searching easy.

### Be nice to machines

Machines usually get confused by spaces, special characters (like `^.*?+|$"`),
and accents. Some operating systems also treat uppercase and lowercase as the
same. Our files should, therefore, always separate words with `_` or `-`
(consistently), only use lowercase letters and numbers, and never use accented
characters.

```txt
# Good
draft01_jalapeno_essay.docx

# Bad
Draft "Jalapeño" Essay(1).docx
```

An important exception here are file extensions: R scripts should be terminated
by `.R`, always uppercase. This is a longstanding tradition, so it takes
precedence over the other rules.

### Be nice to humans

Humans need context, so short and vague file names are to be avoided. A good
practice is to begin the file name with its "type" (e.g. analysis, note,
report, etc.) and then append a meaningful description.

```txt
# Good
analysis01_descriptive-statistics.R
notes02_tentative-write-up.docx

# Bad
01.R
notes2.docx
```

Note how, in the examples above, `-` is used to separate words that belong to
the same "chunk", and `_` is used to separate different "chunks". I don't follow
this suggestion personally, but many people like the idea.

### Make sorting and searching easy

If your files names need to include dates, always follow the `YYYY-MM-DD` format
and place them before the description so that the files are sorted
automatically.

```txt
# Good
2012-01-01.R
2012-01-02.R
2012-04-01.R

# Bad
1-April-2012.R
1-Jan-2012.R
2-Jan-2012.R
```

If you need to order files by something other than date, use numbers as prefix,
but remember to left pad with `0` so that all numbers have the same length. Also
include keywords if you want to divide files by type.

```txt
# Good
reading01_shakespeare_romeo-and-juliet.docx
reading02_shakespeare_romeo-and-juliet.docx
...
reading11_shakespeare_romeo-and-juliet.docx
notes01_shakespeare_romeo-and-juliet.docx
...
```

## Profile

Most users don't know about a little file called `.Rprofile` (the dot at the
beginning makes the file invisible). Your `.Rprofile` contains R code that is run
every time you open R, so this is a good place to set some options and
configurations. It's also an amazing place to make your analyses impossible to
reproduce.

If you've never used this feature of R, I suggest you stay this way until you
are a seasoned veteran in the language. A good tip for using your `.Rprofile`
correctly is to never put anything that you would write in a script there.

Here is a good example of an `.Rprofile`:

```{r}
#| eval: false

# Set repo
options(repos = c(CRAN = "https://cran.rstudio.org"))

# Change width (like in the beginning of the chapter)
if (interactive()) {
  options(width = 10)
}
```

If you _really_ want to add something to your `.Rprofile`, you can install the
[usethis package](https://usethis.r-lib.org/) and run
`usethis::edit_r_profile()`.

## Versions

To me, this is the most important part of the chapter. You can fix all of your
file names and remove everything from your `.Rprofile`; if you don't update
your system, it's all for naught.

R is a relatively old language, so it has been updated
[many times](https://cran.r-project.org/bin/windows/base/old/) since its launch.
Every two to three months, the R Core Team releases a new version with many
improvements and bug fixes. These updates can be very minor, but sometimes there
are truly exciting new features. in R 4.1.0, for example, the native pipe (`|>`)
was introduced, pretty much retiring magrittr.

This means that, if you don't upgrade your R installation, you'll be missing out
on the evolution of the language and leaving your system vulnerable to known
bugs.

The same is true for packages, the life and blood of R. Packages change much
more and much faster than R itself, since they're not always worried about
backwards compatibility. If you leave your packages to rot, expect your scripts
to stop working on other peoples' machines, to have harder time googling bugs,
and to get stuck with old code.

My advice is then the following:

1. Every few weeks make sure to run `update.packages()` to update your packages.

1. Every few months you should also update the language. If you're on Windows
you can use the [installr package](https://talgalili.github.io/installr/),
otherwise you can go to their [website](https://www.r-project.org/) and grab the
newest binary.

It's evidently a bit harder to keep your environment up to date than to just
keep it unchanged forever. Sometimes you'll have to rewrite something that
stopped working from one version to the next, but this is much easier than
letting your system grow more and more out of date until, all of the sudden,
everything stops working all at once. If you keep up with every release, the
adjustments will be small and frequent; if you don't, the adjustments will be
major and you might not be able to make them in a reasonable amount of time.

As a last suggestion, also make sure to update your operating system. R doesn't
support every version of Windows and Mac forever, and I'm not even mentioning
the security patches you're missing when you choose to "update later".
